---
title: "INLA analysis of Latin Square design - one fixed and two crossed random effects"
author: "[Julian Faraway](https://julianfaraway.github.io/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE,comment=NA, fig.path="figs/", warning=FALSE, message=FALSE, optipng='-o7', pngquant='--speed=1 --quality=0-50')
options(digits=5,show.signif.stars=FALSE,width=120)
knit_hooks$set(optipng = hook_optipng)
knit_hooks$set(pngquant = hook_pngquant)
knitr::knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 20)
})
opts_chunk$set(setPch = TRUE)
```

See the [introduction](index.md) for an overview. 
Load the libraries:

```{r}
library(ggplot2)
library(INLA)
```

# Data

```{r abra}
data(abrasion, package="faraway")
abrasion
ggplot(abrasion,aes(x=material, y=wear, shape=run, color=position))+geom_point(position = position_jitter(width=0.1, height=0.0))
```

# Default prior model


```{r}
formula <- wear ~ material + f(run, model="iid") + f(position, model="iid")
result <- inla(formula, family="gaussian", data=abrasion)
result <- inla.hyperpar(result)
summary(result)
```

The run and position precisions look far too high. Need to change the default prior

# Informative Gamma priors on the precisions

Now try more informative gamma priors for the random effect precisions. Define it so
the mean value of gamma prior is set to the inverse of the variance of
the residuals of the fixed-effects only model. We expect the error
variances to be lower than this variance so this is an overestimate.
The variance of the gamma prior (for the precision) is controlled by
the `apar` shape parameter.

```{r}
apar <- 0.5
lmod <- lm(wear ~ material, abrasion)
bpar <- apar*var(residuals(lmod))
lgprior <- list(prec = list(prior="loggamma", param = c(apar,bpar)))
formula = wear ~ material+f(run, model="iid", hyper = lgprior)+f(position, model="iid", hyper = lgprior)
result <- inla(formula, family="gaussian", data=abrasion)
result <- inla.hyperpar(result)
summary(result)
```

Compute the transforms to an SD scale for the random effect terms. Make a table of summary statistics for the posteriors:

```{r sumstats}
sigmarun <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[2]])
sigmapos <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[3]])
sigmaepsilon <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[1]])
restab=sapply(result$marginals.fixed, function(x) inla.zmarginal(x,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmarun,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmapos,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaepsilon,silent=TRUE))
colnames(restab) = c("mu","B-A","C-A","D-A","run","position","epsilon")
data.frame(restab)
```

Also construct a plot the SD posteriors:

```{r plotsdsab}
ddf <- data.frame(rbind(sigmarun,sigmapos,sigmaepsilon),errterm=gl(3,nrow(sigmarun),labels = c("run","position","epsilon")))
ggplot(ddf, aes(x,y, linetype=errterm))+geom_line()+xlab("wear")+ylab("density")+xlim(0,35)
```

Posteriors look OK although no weight given to smaller values.

# Penalized Complexity Prior

In [Simpson et al (2015)](http://arxiv.org/abs/1403.4630v3), penalized complexity priors are proposed. This
requires that we specify a scaling for the SDs of the random effects. We use the SD of the residuals
of the fixed effects only model (what might be called the base model in the paper) to provide this scaling.

```{r}
lmod <- lm(wear ~ material, abrasion)
sdres <- sd(residuals(lmod))
pcprior <- list(prec = list(prior="pc.prec", param = c(3*sdres,0.01)))
formula = wear ~ material+f(run, model="iid", hyper = pcprior)+f(position, model="iid", hyper = pcprior)
result <- inla(formula, family="gaussian", data=abrasion)
result <- inla.hyperpar(result)
summary(result)
```

Compute the summaries as before:

```{r ref.label="sumstats"}
```

Make the plots:

```{r abrapc, ref.label="plotsdsab"}
```

Posteriors put more weight on lower values compared to gamma prior.
Some work is necessary to correctly compute the tails of the densities.

# Package version info

```{r}
sessionInfo()
```
