---
title: INLA analysis of a longitudinal model
author: "[Julian Faraway](http://people.bath.ac.uk/jjf23/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
  toc: true
  theme: cosmo
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,comment=NA, fig.path="/tmp/Figs/", warning=FALSE, message=FALSE)
options(digits=5,show.signif.stars=FALSE,width=120)
```

See the [introduction](index.html) for an overview. Also see [other analyses](../mixchange/multilevel.html) of this data.
See the textbook for a full discussion.

Load the libraries:

```{r}
library(ggplot2)
library(INLA)
library(dplyr)
```

# Data

Load in and plot the data:

```{r}
data(psid, package="faraway")
head(psid)
summary(psid)
psid$cyear <- psid$year-78
psid20 <- filter(psid, person <= 20)
ggplot(psid20, aes(x=year, y=income))+geom_line()+facet_wrap(~ person)
ggplot(psid20, aes(x=year, y=income+100, group=person)) +geom_line()+facet_wrap(~ sex)+scale_y_log10()
```


# Default prior model


```{r}
psid$slperson <- psid$person
formula <- log(income) ~ cyear*sex+age+educ + f(person, model="iid") + f(slperson, cyear , model="iid")
result <- inla(formula, family="gaussian", data=psid)
result <- inla.hyperpar(result)
summary(result)
```

Seems like the default priors might be adequate.
Compute the transforms to an SD scale for the random effect terms. Make a table of summary statistics for the posteriors:

```{r sumstats}
sigmaint <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[2]])
sigmaslope <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[3]])
sigmaepsilon <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[1]])
restab=sapply(result$marginals.fixed, function(x) inla.zmarginal(x,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaint,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaslope,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaepsilon,silent=TRUE))
mm <- model.matrix(~ cyear*sex+age+educ,psid)
colnames(restab) = c(colnames(mm),"intSD","slopeSD","epsilon")
data.frame(restab)
```

Also construct a plot the SD posteriors:

```{r plotsds}
ddf <- data.frame(rbind(sigmaint,sigmaslope,sigmaepsilon),errterm=gl(3,1024,labels = c("slope","intercept","epsilon")))
ggplot(ddf, aes(x,y, linetype=errterm))+geom_line()+xlab("log(income)")+ylab("density")
```

Posteriors look OK. Slope SD is on a different scale so not so comparable.

# Session Info

```{r}
sessionInfo()
```
