---
title: INLA analysis of a Multiple Response Multilevel  model
author: "[Julian Faraway](http://people.bath.ac.uk/jjf23/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
  toc: true
  theme: cosmo
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,comment=NA, fig.path="/tmp/Figs/", warning=FALSE, message=FALSE)
options(digits=5,show.signif.stars=FALSE,width=96)
```

See the [introduction](index.html) for an overview. Also see [other analyses](../mixchange/multiple.html) of this data.
See the textbook for a full discussion. See a [STAN analysis](../stan/multiple.html) of the same data.

Load the libraries:

```{r}
library(ggplot2)
library(INLA)
```

# Data

Load in and summarize the data:

```{r}
data(jsp, package="faraway")
jspr <- jsp[jsp$year==2,]
mjspr <- data.frame(rbind(jspr[,1:6],jspr[,1:6]),subject=factor(rep(c("english","math"),c(953,953))),score=c(jspr$english/100,jspr$math/40))
mjspr$craven <- mjspr$raven-mean(mjspr$raven)

```

Plot the data
```{r}
ggplot(mjspr, aes(x=raven, y=score))+geom_jitter(alpha=0.25)+facet_grid(gender ~ subject)
```



# Default prior model

Need to construct unique labels for nested factor levels of class and student:


```{r}
mjspr$school <- factor(mjspr$school)
mjspr$classch <- factor(paste(mjspr$school,mjspr$class,sep="."))
mjspr$classchid <- factor(paste(mjspr$school,mjspr$class,mjspr$id,sep="."))
```

```{r}
formula <- score ~ subject*gender+craven*subject+social + f(school, model="iid") + f(classch, model="iid") + f(classchid, model="iid")
result <- inla(formula, family="gaussian", data=mjspr)
result <- inla.hyperpar(result)
summary(result)
```

The class precision looks too high. Need to change the default prior

# Informative Gamma priors on the precisions

Now try more informative gamma priors for the precisions. Define it so
the mean value of gamma prior is set to the inverse of the variance of
the residuals of the fixed-effects only model. We expect the error
variances to be lower than this variance so this is an overestimate.
The variance of the gamma prior (for the precision) is controlled by
the `apar` parameter.

```{r}
apar <- 0.5
lmod <- lm(score ~ subject*gender+craven*subject+social,mjspr)
bpar <- apar*var(residuals(lmod))
lgprior <- list(prec = list(prior="loggamma", param = c(apar,bpar)))
formula = score ~ subject*gender+craven*subject+social+f(school, model="iid", hyper = lgprior)+f(classch, model="iid", hyper = lgprior)+f(classchid, model="iid", hyper = lgprior)
result <- inla(formula, family="gaussian", data=mjspr)
result <- inla.hyperpar(result)
summary(result)
```

Compute the transforms to an SD scale for the field and error. Make a table of summary statistics for the posteriors:

```{r sumstats}
sigmaschool <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[2]])
sigmaclass <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[3]])
sigmaid <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[4]])
sigmaepsilon <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[1]])
restab=sapply(result$marginals.fixed, function(x) inla.zmarginal(x,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaschool,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaclass,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaid,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaepsilon,silent=TRUE))
colnames(restab) = c(names(lmod$coef),"school","class","id","epsilon")
data.frame(restab)
```

Also construct a plot of the SD posteriors:

```{r plotsds}
ddf <- data.frame(rbind(sigmaschool,sigmaclass,sigmaid,sigmaepsilon),errterm=gl(4,1024,labels = c("school","class","id","epsilon")))
ggplot(ddf, aes(x,y, linetype=errterm))+geom_line()+xlab("score")+ylab("density")+xlim(0,0.15)
```

Posteriors look OK. More variation is assigned to the school and class than previous models.

# Penalized Complexity Prior

In [Simpson et al (2015)](http://arxiv.org/abs/1403.4630v3), penalized complexity priors are proposed. This
requires that we specify a scaling for the SDs of the random effects. We use the SD of the residuals
of the fixed effects only model (what might be called the base model in the paper) to provide this scaling.

```{r}
lmod <- lm(score ~ subject*gender+craven*subject+social,mjspr)
sdres <- sd(residuals(lmod))
pcprior <- list(prec = list(prior="pc.prec", param = c(3*sdres,0.01)))
formula = score ~ subject*gender+craven*subject+social+f(school, model="iid", hyper = pcprior)+f(classch, model="iid", hyper = pcprior)+f(classchid, model="iid", hyper = pcprior)
result <- inla(formula, family="gaussian", data=mjspr)
result <- inla.hyperpar(result)
summary(result)
```

Compute the summaries as before:

```{r ref.label="sumstats"}
```

Make the plots:

```{r ref.label="plotsds"}
```

Class variation is quite small compared to the other sources.

# Package version info

```{r}
sessionInfo()
```
