---
title: INLA analysis of a multilevel model
author: "[Julian Faraway](http://people.bath.ac.uk/jjf23/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
  toc: true
  theme: cosmo
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,comment=NA, fig.path="/tmp/Figs/", warning=FALSE, message=FALSE)
options(digits=5,show.signif.stars=FALSE,width=96)
```

See the [introduction](index.html) for an overview. Also see [other analyses](../mixchange/multilevel.html) of this data.
See the textbook for a full discussion.

Load the libraries:

```{r}
library(ggplot2)
library(INLA)
```

# Data

Load in and plot the data. We centre the Raven score and create unique labels for the classes within schools:

```{r}
data(jsp, package="faraway")
jspr <- jsp[jsp$year==2,]
jspr$craven <- jspr$raven-mean(jspr$raven)
jspr$classch <- paste(jspr$school,jspr$class,sep=".")
ggplot(jspr, aes(x=raven, y=math))+xlab("Raven Score")+ylab("Math Score")+geom_point(position = position_jitter())
ggplot(jspr, aes(x=social, y=math))+xlab("Social Class")+ylab("Math Score")+geom_boxplot()
```


# Default prior model


```{r}
formula <- math ~ social+craven + f(school, model="iid") + f(classch, model="iid")
result <- inla(formula, family="gaussian", data=jspr)
result <- inla.hyperpar(result)
summary(result)
```

The school precision looks far too high. Need to change the default prior

# Informative Gamma priors on the precisions

Now try more informative gamma priors for the precisions. Define it so
the mean value of gamma prior is set to the inverse of the variance of
the residuals of the fixed-effects only model. We expect the error
variances to be lower than this variance so this is an overestimate.
The variance of the gamma prior (for the precision) is controlled by
the `apar` shape parameter in the code.

```{r}
apar <- 0.5
lmod <- lm(math ~ social+craven, jspr)
bpar <- apar*var(residuals(lmod))
lgprior <- list(prec = list(prior="loggamma", param = c(apar,bpar)))
formula = math ~ social+craven+f(school, model="iid", hyper = lgprior)+f(classch, model="iid", hyper = lgprior)
result <- inla(formula, family="gaussian", data=jspr)
result <- inla.hyperpar(result)
summary(result)
```

Compute the transforms to an SD scale for the random effect terms. Make a table of summary statistics for the posteriors:

```{r sumstats}
sigmaschool <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[2]])
sigmaclass <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[3]])
sigmaepsilon <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[1]])
restab <- sapply(result$marginals.fixed, function(x) inla.zmarginal(x,silent=TRUE))
restab <- cbind(restab, inla.zmarginal(sigmaschool,silent=TRUE))
restab <- cbind(restab, inla.zmarginal(sigmaclass,silent=TRUE))
restab <- cbind(restab, inla.zmarginal(sigmaepsilon,silent=TRUE))
colnames(restab) <- c(names(lmod$coef),"school","class","epsilon")
data.frame(restab)
```

Also construct a plot of the SD posteriors:

```{r plotsds}
ddf <- data.frame(rbind(sigmaschool,sigmaclass,sigmaepsilon),errterm=gl(3,1024,labels = c("school","class","epsilon")))
ggplot(ddf, aes(x,y, linetype=errterm))+geom_line()+xlab("math")+ylab("density")+xlim(0,7)
```

These look reasonable.

# Penalized Complexity Prior

In [Simpson et al (2015)](http://arxiv.org/abs/1403.4630v3), penalized complexity priors are proposed. This
requires that we specify a scaling for the SDs of the random effects. We use the SD of the residuals
of the fixed effects only model (what might be called the base model in the paper) to provide this scaling.

```{r}
lmod <- lm(math ~ social+craven, jspr)
sdres <- sd(residuals(lmod))
pcprior <- list(prec = list(prior="pc.prec", param = c(3*sdres,0.01)))
formula = math ~ social+craven+f(school, model="iid", hyper = pcprior)+f(classch, model="iid", hyper = pcprior)
result <- inla(formula, family="gaussian", data=jspr)
result <- inla.hyperpar(result)
summary(result)
```

Compute the summaries as before:

```{r ref.label="sumstats"}
```

Make the plots:

```{r ref.label="plotsds"}
```

Posteriors look OK.


# Package version info

```{r}
sessionInfo()
```
