---
title: Non-stationary smoothing with Gaussian process regression using INLA
bibliography: merged.bib
author: "[Julian Faraway](https://julianfaraway.github.io/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE,comment=NA, fig.path="figs/", warning=FALSE, message=FALSE, optipng='-o7', pngquant='--speed=1 --quality=0-50')
options(digits=5,show.signif.stars=FALSE,width=120)
knit_hooks$set(optipng = hook_optipng)
knit_hooks$set(pngquant = hook_pngquant)
knitr::knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 20)
})
opts_chunk$set(setPch = TRUE)
```

See the [introduction](index.md) for more about INLA. See an example
of [Gaussian process regression](gpreg.md).
The construction is detailed in 
our [book](http://julianfaraway.github.io/brinla/).

Load the packages (you may need to [install the brinla package](https://github.com/julianfaraway/brinla))

```{r nonstat-1}
library(INLA)
library(brinla)
```

# Data

We simulate some data with a known true function.

```{r nonstat-2}
set.seed(1)
n <- 100
x <- seq(0, 1, length = n)
f.true <- (sin(2 * pi * (x)^3))^3
y <- f.true + rnorm(n, sd = 0.2)
td <- data.frame(y = y, x = x, f.true)
```

and plot it:

```{r nonstat-3}
plot(y ~ x, td)
lines(f.true ~ x, td)
```

This function is challenging to fit because of the varying amount of smoothness. You can
see examples of various smoothing methods applied to this data in [@Faraway:16b].

# GP fitting

The default fit uses priors based on the SD of the response and the range of the predictor
to motivate sensible priors.

```{r nonstat-4}
gpmod = bri.gpr(td$x, td$y)
```

We can plot the resulting fit and 95% credible bands

```{r nonstat-5}
plot(y ~ x, td, col = gray(0.75))
lines(gpmod$xout, gpmod$mean)
lines(gpmod$xout, gpmod$lcb, lty=2)
lines(gpmod$xout, gpmod$ucb, lty=2)
lines(f.true ~ x, td, col=2)
```

On the right end of the function, the fit is too rough. On the left end, it is too smooth. The minimum at 0.9 is
underestimated and the true function is even outside the credibility bands in this region. The problem
is that the standard method uses the same amount of smoothness everywhere.

# Variable smoothness

We can allow the smoothness to vary. A description of this can be found in [@lind:rue:jss:15] and also
in our [book](http://julianfaraway.github.io/brinla/). We have implemented this in our
`brinla` package:

```{r nonstat-6}
fg <- bri.nonstat(td$x, td$y)
plot(y ~ x, td, col = gray(0.75))
lines(f.true ~ x, td, col = 2)
lines(fg$xout, fg$lcb, lty=2)
lines(fg$xout, fg$ucb, lty=2)
lines(fg$xout, fg$mean)
```

We achieve a smoother fit on the left but keep the less smooth fit on the right.

# Package versions

```{r nonstat-7}
sessionInfo()
```



# References







