---
title: "Define your own prior with INLA"
author: "[Julian Faraway](http://people.bath.ac.uk/jjf23/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,comment=NA, fig.path="/tmp/Figs/", warning=FALSE, message=FALSE)
options(digits=5,show.signif.stars=FALSE,width=120)
```

## Initialization

See the [introduction](index.html) for an overview.
Load the libraries:

```{r}
library(INLA)
library(brinla)
library(ggplot2)
```

We illustrate the method using the following data:

```{r}
data(reeds, package="brinla")
ggplot(reeds,aes(site,nitrogen))+geom_point()
```

We will build a model of the form:
```
nitrogen = intercept + site + error
```

where `intercept` is a fixed effect modeled with a normal prior with a zero mean and large variance.
The `site` is a random effect and `error` is the error term. These have distributions which are
normal with mean zero and some variance. These variances are treated as hyperparameters with
priors we shall specify.

# Analysis with default prior

By default, INLA places a gamma prior on the precisions (inverse of the variance). We
fit this model and examine the posteriors in terms of the SDs (which are easier to understand than precisions)

```{r}
formula <- nitrogen ~ 1 + f(site, model="iid")
imod <- inla(formula,family="gaussian", data = reeds)
bri.hyperpar.summary(imod)
```

We can also plot the posteriors of these two SDs.

```{r}
bri.hyperpar.plot(imod)
```


# PC prior

INLA has [other choices](http://www.r-inla.org/models/priors) for the priors. Here
is an example using the [penalized complexity prior](https://arxiv.org/abs/1403.4630). We use the SD of the response to
help us set the scale of this prior (although it is better if you set this with knowledge
of the problem behind the data)

```{r}
sdres <- sd(reeds$nitrogen)
pcprior <- list(prec = list(prior="pc.prec", param = c(3*sdres,0.01)))
formula <- nitrogen ~ f(site, model="iid", hyper = pcprior)
pmod <- inla(formula, family="gaussian", data=reeds)
bri.hyperpar.summary(pmod)
```

Plot the posterior densities.

```{r}
bri.hyperpar.plot(pmod)
```


# Half Cauchy prior

It is possible to set our own prior for the SD of the `site` effect. The use of the
half(ie. positive part of) Cauchy is a commonly used choice which is not directly programmed
in INLA. The density for the SD ($\sigma$) with scaling $\lambda$ is:
$$
 p(\sigma | \lambda) = \frac{2}{\pi\lambda(1+(\sigma/\lambda)^2)}, \quad\quad \sigma \ge 0.
$$

We calculate scaling equivalent to the PC prior scaling for future use:

```{r}
(lambda = 3*sdres/tan(pi*0.99/2))
```

INLA works with the precision and the calculation requires the log density of the precision, $\tau$, which is
$$
  \log p(\tau | \lambda) = -\frac{3}{2}\log\tau - \log (\pi\lambda) - \log (1+1/(\tau\lambda^2))
$$


We then define the prior as:


```{r}
halfcauchy = "expression:
              lambda = 0.022;
              precision = exp(log_precision);
              logdens = -1.5*log_precision-log(pi*lambda)-log(1+1/(precision*lambda^2));
              log_jacobian = log_precision;
              return(logdens+log_jacobian);"
```


And use this prior with

```{r}
hcprior = list(prec = list(prior = halfcauchy))
formula <- nitrogen ~ f(site, model="iid", hyper = hcprior)
hmod <- inla(formula, family="gaussian", data=reeds)
bri.hyperpar.summary(hmod)
```

Plot the posterior distributions:

```{r}
bri.hyperpar.plot(hmod)
```

The result is quite similar to the penalized complexity prior.



